Towards a Complete Model of Mutation Selection Balance in Humans

### 2023-05-12 manuscript/manuscript.tex
more theory section stuff

Both the classic background selection and SC16 models imagine a negative
selection process where a continual influx of deleterious mutations flow into
the population at a rate of $\mu$ per basepair per generation in a conserved
region of $L$ basepairs, such that the region-wide per generation mutation rate
is $U = \mu L$. Each mutation imposes an additive selective cost of $s$ in
heterozygotes and $2s$ in homozygotes. The classic background selection model
considers when mutations are sufficiently deleterious that they are destined to
extinction
\parencite{Charlesworth1993-gb,Nordborg1996-nq,Hudson1995-pt,Hudson1994-oh}. In
this case, the genealogy is well-approximated by a neutral coalescence process
with a rescaled effective population size of $N_0 = N \exp(-\nicefrac{U}{s})$.
This approximation works reasonably well for two reasons (though see
\textcite{Cvijovic2018-vd}). First, when selection is strong, the number of
deleterious mutations per haplotype has a stationary Poisson distribution with
mean equal to the equilibrium under the deterministic mutation-selection model
\parencite{Haldane1927-ga}. Since lineages carrying any strongly deleterious
mutations cannot contribute to the genealogy, coalescence events only occur in
the fraction $\exp(-\nicefrac{U}{s}$ free of mutations. Second, two present-day
lineages carrying any deleterious mutations trace their ancestry back to the
pool of $N_0$ mutation-free ancestors very quickly, since these harmful
mutations could only have been acquired recently by the lineage. On coalescence
timescales of $\sim 2N_0$, this time is negligible and can be ignored
(\textcite{Durrett2008-ql}, p. 213; \textcite{Good2014-yz}).

classic background selection model are only applicable for strongly deleterious
mutations that have no probability of fixation
\parencite{Good2014-yz,Gordo2002-dr,Charlesworth2013-kl}.

Modifications of this basic model that allow for recombination between the
segment under negative selection and a focal neutral site
\parencite{Nordborg1996-nq,Hudson1994-oh} have been the basis of statistical
methods that fit genome-wide patterns of diversity to theoretic expectations
\parencite{McVicker2009-ax,Hudson1995-xc}. Furthermore, methods that
incorporate both hitchhiking and background selection allow for estimates of
the relative contributions of these two modes of selection and model
comparisions \parencite{Elyashiv2016-vt,Murphy2022-sj}.

However, the predicted reduction in diversity under the classic background
selection model is inaccurate whenever selection is weak enough that
deleterious mutations may rise in frequency in the population and possibly
reach fixation. This is a well-known problem (XXX), that is mathematically
difficult because the

>>>>

Both the classic background selection and SC16 models imagine a negative
selection process where a continual influx of deleterious mutations flow into
the population at a rate of $\mu$ per basepair per generation in a conserved
region of $L$ basepairs, such that the region-wide per generation mutation rate
is $U = \mu L$. Each mutation imposes an additive selective cost of $s$ in
heterozygotes and $2s$ in homozygotes. The classic background selection model
considers when mutations are sufficiently deleterious that they are destined to
extinction
\parencite{Charlesworth1993-gb,Nordborg1996-nq,Hudson1995-pt,Hudson1994-oh}. In
this case, the genealogy is well-approximated by a neutral coalescence process
with a rescaled effective population size of $N_0 = N \exp(-\nicefrac{U}{s})$.
This approximation works reasonably well for two reasons (though see
\textcite{Cvijovic2018-vd,Walczak2012-fi,Nicolaisen2012-vs}). First, when
selection is strong, the number of deleterious mutations per haplotype has a
stationary Poisson distribution with an average equal to the equilibrium under
the deterministic mutation-selection model \parencite{Haldane1927-ga}. Since
lineages carrying any strongly deleterious mutations cannot contribute to the
genealogy, coalescence events only occur in the fraction
$\exp(-\nicefrac{U}{s}$ of the population free of mutations. Second, the time
it takes for a present-day lineage to trace their ancestry back to a
mutation-free ancestor (the ``delay phase") is negligible compared to the
coalescence timescale of $N_0$ among these ancestors (the ``coalescence
phase"), and can be ignored (\textcite{Durrett2008-ql}, p. 213;
\textcite{Good2014-yz}).

When mutations are weakly deleterious, their frequency trajectories are no
longer well-approximated by deterministic models. This is because the selection
on an allele is only weak relative to the scale of stochastic perturbations it
encounters along its trajectory. When genetic drift is the only source of
randomness in allele frequency change, the relevant scale of stochastic
perturbations is the drift-effective population size $N$ and selection dynamics
are governed by compound parameter $2Ns$. When $2Ns \gg 1$, the stochastic
component of allele frequency change is minuscule relative to selection and can
be ignored as the classic BGS theory does; this is why the rescaling factor
under the BGS model does not depend on $N$. However, when $2Ns \le 1$
stochastic perturbations are large relative to selection, and a weakly
deleterious allele can drift up to intermediate frequencies before eventual
loss or fixation. Weakly deleterious alleles can persist in the population much
longer, and consequently

When this occurs, the distribution of the number of deleterious mutations (and
its corresponding fitness distribution) are no longer stationary and become a
traveling wave \parencite{Rouzine2008-qz,Good2013-lp}. Furthermore, individuals
in the least-loaded class are stochastically lost, fixing those mutations with
a click of ``Muller's ratchet" \parencite{Muller1964-ki}. Determining both the
rate of Muller's ratchet \parencite{Haigh1978-gt,Gordo2002-dr,Gessler1995-hz}
and the shape of genealogies under weak selection
\parencite{Przeworski1999-mb,OFallon2010-my,Higgs1995-xc} have been stubborn
problems
